Ecological Limits to Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning

### 2022-05-13 manuscript/manuscript.Rmd

Forecasting plays an important and rapidly growing role in both testing fundamental understanding of ecological processes and informing ecological applications and conservation decision-making [@dietze_iterative_2018; @schindler_portfolio_2015].
Meanwhile, rapid advances in machine learning have rapidly improved the prevalence and accuracy of short term forecasts in many fields [@something]. 
Will these emerging methods improve the capacity for forecasts in ecological systems as well?
Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Oveskienen2010].
These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are simultaneously more important and more difficult to predict than more gradual changes.
Here, we examine several of the best-known examples of critical transitions in ecological systems.
We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.

@@

<!-- some more verbose bits about bifurcation models probably belongs in discussion instead -->
We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasatoid-host model [@nicholson_balance_1935]. 
Though these models may appear simple when measured against high-dimensional and parameter rich models found in some management contexts such as fisheries, they can exhibit rich nonlinear dynamics and provide greater capacity to generalize [@Levins1966; @Getz2018]. 
These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points which has had important theoretical and practical management outcomes [@Folke2004; @Fischer2009; @Polasky2011].
May's model exhibits alternative stable states. 
In this one-dimensional model, transitions between these states can occur due to intrinsic stochasticity, external forcing, or the gradual environmental change that results in a catastrophic saddle-node bifurcation and generates hysteresis. 
The Nicholson-Bailey model is a two species model which contains a Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.

Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is particularly important question for ecological systems and global environmental change problems.
Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analog in the historical data to date. 
Forecast skill under such no-analog conditions may be particularly relevant to ecological forecasting in the context of global change [@Williams2007].

We provide fully reproducible coded examples in R and python for performing, scoring, and visualizing each of the forecasts considered here.
After considerable time considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting. 
Most of our forecasts use the `darts` framework, a sophisticated and well documented python library with support for a wide range of methods.
Our model-based MCMC forecasts use the `greta` framework, an R library that uses python-based `tensorflow` probability to achieve better performance.
While python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)

@@

We expect greater convergence between methods available in R and python in the future, as already illustrated in the example of `greta`.
Complete code for all examples presented here can be found at https://github.com/boettiger-lab/mee_tipping_point_forecasting.

@@

A yet more difficult forecasting scenario is created by the saddle-node bifurcation.
May's consumer-resource model is a one-dimensional model describing the growth of a 'resource' population (e.g. herbivore) which is grazed by a consumer [@may_population_1979].
As in the Nicholson-Bailey model, in the absence of that predation the resource population density grows under a density-dependent pattern described by a logistic function.
The resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time). 
For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range of ecological and socio-ecological systems [@scheffer_catastrophic_2001].

@@

In this context, predicting the probability of a transition in the future based solely on observations prior to a transition occurring is essentially impossible without additional information constraining the model estimate, as such data is equally consistent with infinitely many models or parameter choices which share the same local linearization about the stable point.  
Unlike the saddle-node bifurcation, there is no slowly warping potential basin which can be detected to inform estimates. 
Thus, in this scenario, rather than considering the problem of predicting the future evolution of a single time series based only on it's historical values, we consider an alternative framing of the task:
we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
Based on this data, our forecaster seeks to identify the distribution of expected transition times for analagous systems starting from the same initial condition. 
This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [@Scheffer2001; @Folke2004].
(Note that such stochastic transitions between alternative stable states can also create oscillatory-like dynamics when stochasticity is sufficiently high enough to drive repeated transitions from one attractor to the other and back again.  In such cases, it might be reasonable to estimate a strictly forward-looking forecast of a single system, predicting the distribution of these transitions.)
Model definition is the same as May's model for the saddle node with fixed parameter $h$, values: $r = 1$, $K = 1$, $s = 0.1$, $h_0 = h = 0.26$, $\alpha = 0$, $\sigma = 0.02$, $N_0 = 0.55$.

@@

We present forecasts produced by ARIMA models as the model-free analogs to the forecasts made using parameter estimation with MCMC.
Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@hyndman_forecasting_2018], these models are not well-suited for problems with complex bifurcation dynamics.
Thus, ARIMA-based forecasts should be treated as a lower bound for the performance of non-mechanistic models.
In contrast to inference with MCMC, uncertainty with ARIMA models is estimated directly from the learned parameters [@hyndman_forecasting_2018].
Since ARIMA is a very commonly encountered forecasting method, we will refer readers to @hyndman_forecasting_2018 for further discussion.

@@

A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@madhyastha_model_2019].
We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@madhyastha_model_2019]. 
Another likely cause is that machine learning models commonly overfit the training data [@mehta_high-bias_2019].
Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized. 
With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@mehta_high-bias_2019].
One frequently cited method to reduce overfitting is K-fold cross validation, but this approach cannot be effectively employed when there is one or few time series in the training set.
To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
We found this simple ensemble method to be an effective way to improve generalizability in the limited data regime.

@@

The Transformer is a state of the art ML architecture that is able to model long and short term dependencies on sequence to sequence tasks [@vaswani_attention_2017].
Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani_attention_2017].
For example, if given a sentence, a transformer could learn the contextual relationship between a subject and a direct object, but a recurrent network would process all the words as one phrase.
Because of self-attention, Transformers do not need to process data sequentially, and thus can be parallelized, offering significant computational advantages [@vaswani_attention_2017].
The Transformer is likely to be a foundational method for future AI research [@bommasani_opportunities_2021], so we considered it critical to investigate Transformers in this study.

@@

We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, stochastic transition).
In addition to these cases, we also consider an "ensemble model", generated from drawing from the distribution of all models except the MCMC model. 
Such ensemble techniques can better reflect uncertainty than relying on any single method [@something_about_ensembles].
For simplicity, we consider the un-weighted case, where each model is represented equally in the ensemble.
Using model-based simulations allows us to examine performance against an multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance. 
By taking the true model structure as given, MCMC methods can be used to determine a theoretical limit of forecasting skill.
Note that in both bifurcation scenarios, future dynamics will see states never previously observed in the historical data that was used to train each of the methods (i.e. very small population sizes).  
This no-analog aspect of forecasting bifurcation dynamics means that even with many sample points in the training data _and_ perfect knowledge of the true model structure, posterior distributions of parameter values are still influenced by the choice of priors.

@@

Forecasts of the Hopf bifurcation (Figure 3) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
All models are trained using 100 time points drawn from the period of time prior to the onset of the Hopf bifurcation, which leads to a stable limit cycle that gradually grows in magnitude.
Most models predict a roughly constant mean with a spread roughly equal to that created by the stochastic oscillations around the stable node as seen in the training data prior to the bifurcation.
Notably, the GRU method picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
However, like the ML models, it fails to predict the growing amplitude of those oscillations.
Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
Despite this, overall all methods score comparably in CRPS score (Fig 2), with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
A clearer picture can be seen by looking at these skill scores over time (Fig 6-7), which show that MCMC is initially performing worse (over-predicting variance) but as oscillations grow further, it starts outperforming the more stationary forecasts of the ML models.

>>>>

Forecasting plays an important and rapidly growing role in both testing our fundamental understanding of ecological processes, and informing ecological applications and conservation decision-making [@dietze_iterative_2018; @schindler_portfolio_2015].
Meanwhile, recent advances in machine learning have rapidly improved the prevalence and accuracy of short term forecasts in many fields [@kao_exploring_2020; @lyu_lstm_2020; @du_multivariate_2020]. 
Will these emerging methods improve the capacity for forecasts in ecological systems as well?
Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Oveskienen2010].
These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are more important and more difficult to predict than gradual changes.
Here, we examine several of the best-known examples of critical transitions in ecological systems.
We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.

@@

<!-- some more verbose bits about bifurcation models probably belongs in discussion instead -->
We will focus the analysis on several different forecasting scenarios based around two classic models in population ecology: 
Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasitoid-host model [@nicholson_balance_1935]. 
Though these models may appear simple when measured against high-dimensional and parameter rich models found in some management contexts such as fisheries, they can exhibit rich nonlinear dynamics and provide greater capacity to generalize [@Levins1966; @Getz2018]. 
These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points which has had important theoretical and practical management outcomes [@Folke2004; @Fischer2009; @Polasky2011].
May's model exhibits alternative stable states. 
In this one-dimensional model, transitions between these states can occur due to intrinsic stochasticity, external forcing, or the gradual environmental change that results in a catastrophic saddle-node bifurcation and generates hysteresis. 
The Nicholson-Bailey model is a two species model which contains a Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.

Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is a particularly important question for ecological systems and global environmental change problems.
Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analog in the historical data to date. 
Forecast skill under such no-analog conditions may be particularly relevant to ecological forecasting in the context of global change [@Williams2007].

We provide fully reproducible coded examples in R and Python for performing, scoring, and visualizing each of the forecasts considered here.
After significant time spent considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting. 
Most of our forecasts use the `darts` framework, a sophisticated and well documented Python library with support for a wide range of methods.
Our model-based MCMC forecasts use the `greta` framework, an R library that uses Python-based `tensorflow` probability to achieve better performance.
While Python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)

@@

We expect greater convergence between methods available in R and Python in the future, as already illustrated in the example of `greta`.
Complete code for all examples presented here can be found at https://github.com/boettiger-lab/mee_tipping_point_forecasting.

@@

A yet more difficult forecasting scenario is created by the saddle-node bifurcation.
May's consumer-resource model is an one-dimensional model describing the growth of a 'resource' population (e.g. herbivore) which is grazed by a consumer [@may_population_1979].
As in the Nicholson-Bailey model, in the absence of that predation, the resource population density grows under a density-dependent pattern described by a logistic function.
The resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time). 
For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range of ecological and socio-ecological systems [@scheffer_catastrophic_2001].

@@

In this context, predicting the probability of a transition in the future based solely on observations prior to a transition occurring is essentially impossible without additional information constraining the model estimate, as such data is equally consistent with infinitely many models or parameter choices which share the same local linearization about the stable point.  
Unlike the saddle-node bifurcation, there is no slowly warping potential basin which can be detected to inform estimates. 
Thus, in this scenario, rather than considering the problem of predicting the future evolution of a single time series based only on its historical values, we consider an alternative framing of the task:
we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
Based on this data, our forecaster seeks to identify the distribution of expected transition times for analagous systems starting from the same initial condition. 
This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [@Scheffer2001; @Folke2004].
(Note that such stochastic transitions between alternative stable states can also create oscillatory-like dynamics when stochasticity is sufficiently high enough to drive repeated transitions from one attractor to the other and back again.  In such cases, it might be reasonable to estimate a strictly forward-looking forecast of a single system, predicting the distribution of these transitions.)
Model definition is the same as May's model for the saddle node with fixed parameter $h$, values: $r = 1$, $K = 1$, $s = 0.1$, $h_0 = h = 0.26$, $\alpha = 0$, $\sigma = 0.02$, $N_0 = 0.55$.

@@

We present forecasts produced by ARIMA models as the model-free analogs to the forecasts made using parameter estimation with MCMC.
Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@hyndman_forecasting_2018], these models are not well-suited for problems with complex bifurcation dynamics.
Thus, ARIMA-based forecasts should be treated as a lower bound for the performance of non-mechanistic models.
In contrast to inference with MCMC, uncertainty with ARIMA models is estimated directly from the learned parameters [@hyndman_forecasting_2018].
Since ARIMA is a commonly encountered method, we will refer readers to @hyndman_forecasting_2018 for further discussion.

@@

A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@madhyastha_model_2019].
We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@madhyastha_model_2019]. 
Another likely cause is that machine learning models commonly overfit the training data [@mehta_high-bias_2019].
Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized [@dar_farewell_2021]. 
With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@mehta_high-bias_2019].
One frequently used method to reduce overfitting is K-fold cross validation [@raschka_model_2020], but this approach cannot be effectively employed when there is one or few time series in the training set.
To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
We found this simple ensemble method to be an effective way to improve generalizability in the limited data regime.

@@

The Transformer is a state of the art ML architecture that is able to model long and short term dependencies on sequence to sequence tasks [@vaswani_attention_2017].
Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani_attention_2017].
For example, if given a sentence, a transformer could learn the contextual relationship between a subject and a direct object, but a recurrent neural network would process all the words as one phrase.
Because of self-attention, Transformers do not need to process data sequentially and thus can be parallelized, offering significant computational advantages [@vaswani_attention_2017].
The Transformer is likely to be a foundational method for future AI research [@bommasani_opportunities_2021], so we considered it critical to investigate Transformers in this study.

@@

We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, stochastic transition).
In addition to these cases, we also consider an "ensemble model", generated by drawing from the distribution of all models except the MCMC model. 
Such ensemble techniques can better reflect uncertainty than relying on any single method [@gneiting_weather_2005].
For simplicity, we consider the unweighted case, where each model is represented equally in the ensemble.
Using model-based simulations allows us to examine performance against a multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance. 
By taking the true model structure as given, MCMC methods can be used to determine a theoretical limit of forecasting skill.
Note that in both bifurcation scenarios, future dynamics will see states never previously observed in the historical data that was used to train each of the methods (i.e. very small population sizes).  
This no-analog aspect of forecasting bifurcation dynamics means that even with many sample points in the training data _and_ perfect knowledge of the true model structure, posterior distributions of parameter values are still influenced by the choice of priors.

@@

Forecasts of the Hopf bifurcation (Figure 3) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
All models are trained using 100 time points drawn from the period of time prior to the onset of the Hopf bifurcation, which leads to a stable limit cycle that gradually grows in magnitude.
Most models predict a roughly constant mean with a spread roughly equal to that created by the stochastic oscillations around the stable node as seen in the training data prior to the bifurcation.
Notably, the GRU method picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
However, like the other ML models, it fails to predict the growing amplitude of those oscillations.
Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
Despite this, overall all methods score comparably in CRPS score (Fig 2), with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
A clearer picture can be seen by looking at these skill scores over time (Fig 6-7), which show that MCMC is initially performing worse (over-predicting variance) but as oscillations grow further, it starts outperforming the more stationary forecasts of the ML models.
